---
title: "PIT Homelessness County Estimates"
author: "UW Epi NCW Consorium Team"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    toc: true
---

This RMarkdown file pulls *5 year American Community Survey (ACS) estimates from the US Census Bureau* using the `tidycensus` R package. This will allow us to standardize the Point In Time (PIT) Count Homelessness data to number of people experiencing homelessness per 100,000 individuals per year.  


# Pull in necessary packages and set `tidycensus` API 

For this step, **you will need to establish your own `tidycensus` API**. You can find this information here: https://walker-data.com/tidycensus/articles/basic-usage.html 

To get started working with `tidycensus`, users should load the package along with the `tidyverse` package, and **set their Census API key.** A key can be obtained from http://api.census.gov/data/key_signup.html

The variable we are interested in is `B01003_001`, which is the estimated "Total Population" in a given geography. However, for every new year, check that this is still the correct variable (see below).

**The updates that will be needed every year are...(CTRT+F "UPDATE NEEDED" to find where updates may be needed)**

1) Adding your own Census API key 
2) Update the years of data you are interested in pulling. This can be found in the first "UPDATE NEEDED" section. You will just add your new year to the `years` vector (see *UPDATE NEEDED* in first code chunk). 
3) This may not need to be updated, but it's worth adding your newest year to the load_variables() function. This is where you pull in the variables you are interested in, and will allow you to check if the "Total Population" variable has changed. It should be `B01003_001`, and is as of the 2022 ACS 5-year estimates.
4) You will need to update the name of the "homelessness_data" every year ONLY if you change the name. 


```{r, include=TRUE, message=FALSE, warning=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)

### Uncomment and install packages as needed 
#install.packages("tidyr")
#install.packages("gmodels")
#install.packages("tidyverse")
#install.packages("tidycensus")
#install.packages("data.table")
#install.packages("table1")
#install.packages("tibble")
#install.packages("purrr")
#install.packages("labelled")
#install.packages("grid")
#install.packages("gridExtra")
#install.packages("janitor")
#install.packages("censusapi")

library(dplyr)
library(tidyverse)
library(tidycensus)
library(data.table)
library(table1)
library(tibble)
library(purrr)
library(labelled)
library(grid)
library(gridExtra)
library(janitor)
library(censusapi)

# Set API (https://api.census.gov/data/key_signup.html) 
census_api_key(key = "UPDATE NEEDED - SET YOUR CENSUS API KEY HERE", overwrite=TRUE, install=TRUE)
readRenviron("~/.Renviron")


# UPDATE NEEDEDL: pull in ACS datasets 
acs_vars_2018 <- load_variables(2018, "acs5", cache = TRUE)
acs_vars_2019 <- load_variables(2019, "acs5", cache = TRUE)
acs_vars_2019 <- load_variables(2019, "acs5", cache = TRUE)
acs_vars_2021 <- load_variables(2021, "acs5", cache = TRUE)
acs_vars_2022 <- load_variables(2022, "acs5", cache = TRUE)


# Set years we're interested in pulling data for 
# As of 4/25/2024, 2023 5 year ACS estimates are not available
# UPDATE NEEDED: this is where you will need to add the new year of data you are interested in pulling
# UPDATE NEEDED (potentially): the B01003_001 variable, aka the "Total Population" variable, is current as of 2022 ACS 5 year estimates. This is unlikely to change, but worth checking 
years <- c(2018, 2019, 2020, 2021, 2022)

# Set homelessness data name and pull it into your environment
# UPDATE NEEDED: change the CSV name if it has been updated 
homelessness_data <- read.csv("homelessness_data.csv")



```

# Washington State

```{r, include=TRUE, message=FALSE, warning=FALSE}
# Create an empty list to store the datasets
WA_state_pop_data <- list()

# Loop through the years in the vector above, and then save each dataset in environment
for (i in years) {
  WA_state_pop_data[[as.character(i)]] <- get_acs(geography = "state",
                                            variables = c("B01003_001"), 
                                            state = "WA",
                                            survey = "acs5",
                                            year = i)
  
  WA_state_pop_data[[as.character(i)]] <- WA_state_pop_data[[as.character(i)]] %>% 
    dplyr::select(-moe, -GEOID, -NAME) %>% 
    pivot_wider(names_from = variable, values_from = estimate) %>% 
    rename(totalpop = B01003_001) %>% 
    mutate(year = i, location = "WA")
    
  assign(paste0("WA_pop_", i), WA_state_pop_data[[as.character(i)]], envir = .GlobalEnv)
  
}

# Combine all WA state datasets 
WA_state_pop <- bind_rows(WA_state_pop_data)


```



# Counties 

```{r, include=TRUE, message=FALSE, warning=FALSE}
# Create an empty list to store the datasets
WA_county_pop_data <- list()

# Loop through the years in the vector above, restrict to our counties of interest, and then save each dataset in environment
for (i in years) {
  # Fetch ACS data
  WA_county_pop_data[[as.character(i)]] <- get_acs(geography = "county", 
                         variables = c("B01003_001"), 
                         state = "WA", 
                         survey = "acs5", 
                         year = i) %>% 
    filter(NAME %in% c("Chelan County, Washington", 
                       "Douglas County, Washington", 
                       "Grant County, Washington", 
                       "Kittitas County, Washington", 
                       "Okanogan County, Washington"))
  
  # Process and clean the data
  WA_county_pop_data[[as.character(i)]] <- WA_county_pop_data[[as.character(i)]] %>% 
    dplyr::select(-moe, -GEOID) %>% 
    pivot_wider(names_from = variable, values_from = estimate) %>% 
    rename(totalpop = B01003_001, location = NAME) %>%  
    mutate(year = i, 
           location = case_when(
             location == "Chelan County, Washington"   ~ "Chelan",
             location == "Douglas County, Washington"  ~ "Douglas",
             location == "Grant County, Washington"    ~ "Grant",
             location == "Kittitas County, Washington" ~ "Kittitas",
             location == "Okanogan County, Washington" ~ "Okanogan"))
  
  # Create a dataset that's just Chelan and Douglas counties for a given year
  chelandouglas_data <- WA_county_pop_data[[as.character(i)]] %>%
    filter(location %in% c("Chelan", "Douglas")) %>%
    mutate(totalpop = sum(totalpop), location = "Chelan-Douglas") %>%
    distinct()
  
  # Combined dataset for Chelan and Douglas counties with the rest of the counties
  WA_county_pop_data[[as.character(i)]] <- rbind(WA_county_pop_data[[as.character(i)]], chelandouglas_data) %>%
    filter(location != "Chelan" & location != "Douglas")
  
  # Assign the result to the global environment
  assign(paste0("WA_county_pop_", i), WA_county_pop_data[[as.character(i)]], envir = .GlobalEnv)
    
}

# Combine all WA county datasets
WA_county_pop <- bind_rows(WA_county_pop_data)


```


# Combine county and state estimates 

```{r, include=TRUE, message=FALSE, warning=FALSE}
# Reorder columns so they can be combined 
WA_state_pop  <- WA_state_pop  %>% select(year, location, totalpop)
WA_county_pop <- WA_county_pop %>% select(year, location, totalpop)


# Stack dataframes 
WA_pop_PITcount_data <- 
  rbind(WA_state_pop, WA_county_pop) %>% 
  arrange(year, location)


# Slight renames to easily merge with PIT data
WA_pop_PITcount_data <- WA_pop_PITcount_data %>% 
  rename(Year = year, Location = location)


# Save the data in CSV format
write.csv(WA_pop_PITcount_data, "WA_pop_PITcount_data.csv", row.names = FALSE)


# Combine WA_pop_PITcount_data with homelessness data 
homelessness_data$Value <- as.numeric(gsub(",", "", homelessness_data$Value))
homelessness_data_popest <- left_join(homelessness_data, WA_pop_PITcount_data, by = c("Year", "Location"))


# If missing totalpop information, pull from the most recent year with data for the corresponding location 
homelessness_data_popest <- homelessness_data_popest %>% 
  group_by(Location) %>% 
  mutate(totalpop = ifelse(is.na(totalpop), lag(totalpop, order_by = Year), totalpop)) %>% 
  ungroup()


# Create a new column for the standardized rate of homelessness per 100,000 and 10,000 individuals then round the two variables we made
homelessness_data_popest <- homelessness_data_popest %>% 
  mutate(homelessness_rate_100k = (Value / totalpop) * 100000, 
         homelessness_rate_100k = round(homelessness_rate_100k, 2),
         homelessness_rate_10k  = (Value / totalpop) * 10000,
         homelessness_rate_10k  = round(homelessness_rate_10k, 2))


# Save the data in CSV format
write.csv(homelessness_data_popest, "homelessness_data_popest.csv", row.names = FALSE)


```










